What should I notice? Evaluating memorability of events to generate surprising causal candidates.

### 2021-07-26 paper.tex
Working on the document

\begin{abstract}
  When confronted to an unprecedented situation, humans typically show good
  performance in quickly identifying noticeable past events and proposing them
  as possible causal hypotheses. This kind of abductive inference is widely
  overlooked in modern AI approaches which rely on massive datasets to learn
  associative patterns. Our proposal is to formalize and compute a ``memorability''
  score over a memory of various recorded events from a cyber-physical system.
 This score can later be used either to select only relevant information to be
 remembered, or to propose causal hypotheses in unusual situations, on demand.
 As such, the approach aims at being complementary to more traditional
 learning-focused techinques. We provide theoretical ground for our approach by
 using and extending existing results and ideas from Algorithmic Information
 Theory and provide an implementation example, showing practical results in a
 smart-home scenario.
\end{abstract}

@@

The ability to infer new causal relations is one of the many tasks in which
humans prevail but machines still fall short of completing. This ability is
prime in making discoveries, or acquiring new knowledge by making educated
guesses.

Without any knowledge about how physics or electricty work, it would be possible
to infer a possible link between a memorable thunderstorm and a general
black-out only by using the noticeability score of the first. While this kind of
abductive inference can yield many false-positives, its application can be to
propose new hypotheses to be tested out when other possible methods of inference
fail to provide any hypothesis (due to a lack of prior similar situations, or to
the importance of the context). The difficulty of this approach to reasoning is,
however, to quickly identify the relevant candidate hypotheses. In this regards,
we may use a basic instinct: without prior knowledge, the most relevant
hypothesis might simply be the most memorable recent event. However this
definition is highly subjective, and does not seem to fit well into the canvas
of computing and AI.

The difficulty of this computation comes from different factors. First, events
can be of different nature, and not directly comparable. In fact, for the use
case of smart homes, or other cyber physical systems, events range from device
removal to a detected presented or an unusually high temperature. How can one
can then assess which ones are more memorable? Furhtermore, even for comparable
events, different characterics can be put forward to argue for the most
memorable: is a record-high temperature 47 days ago more memorable than the
small deviation recorded just 3 minutes ago? To this day, no current system
proposes to use all this information from various event types from different
devices to compute a unified metrics of ``memorability''.

% The difficulty of doing so comes from the wide variety of events, parameters and
% characteristics to take into account. How can one assess if a small recent
% perturbation is more important than a larger but older one? In addition, one
% needs a common metrics to evaluate the relative importance of data coming from
% various devices. In the case of a smart home, for instance, devices range from
% presence sensors to humidity and temperature sensors, power meters, etc. To this
% day, no current system proposes to analyze all events coming from such various
% devices and come to a unified metrics of ``memorability''.
%

To tackle this challenge, we propose to refer to events using a technique used
by humans: naming. It appears that humans are able to name events and describe
them, regardless of their nature or their characteristics. Furthermore, there is
a noticeable correlation between the lenght of the description, in terms of the
complexity of concepts and arguments, and the ``memorability'' of the event.
Think, for instance, of how ``last year's hottest day'' appears simpler than
``the 182th day of 7 years ago''. Using this method, the day described by the
first sentence would be deemed simpler than the target of the second
description. By comparing the description length of the event with the average
description lenght of similar events, one would then have a measure of the
``memorability'' of events, as memorable events stands out as much simpler (or
more complex) than their neighbours.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{figures/filters}
  \caption{An example of retrieving an event through successive predicative
    filters. From the base memory (yellow), successive filters select the events
    satisfying their predicates (grey arrows). The operations yields successive
    subsets of the memory (yellow), until a final singleton memory is reach,
    meaning that the successive filters retrieve a unique event from the memory.
    The predicates used for this retrival path and their arguments,
    $\pi_{1}, \pi_{2}, \pi_{3}, k_{1}, k_{2}, k_{3}$, can be used as a retrieval
    program. Its length gives an upper bound to the description complexity of
    the event $e_{h}$: ``last year's hottest day''.}
  \label{fig:filters}
\end{figure}

To programatically achieve what humans are easily capable to do, we rely on a
basic model intuition of how memory works. Figure \ref{fig:filters} shows this
process for the example description ``last year's hottest day''. We assume that
a memory $\mathcal{M}$ is a unordered set of various distinct objects, upon
which can be applied boolean predicates $\pi$ to describe them. These predicates thus
provide filters $f_{\pi}$ that can be applied to the memory, selecting all events
satifying predicate $\pi$. By applying successive
filters, we narrow down the size of the memory, until a single event remains.
The complexity of this remaining events can thus be estimated as the simplest
combination of predicative filters used to retrieve it.

The rest of this paper is organized as follows: we first briefly introduce some notions of
Algorithmic Information Theory and its applications in section~\ref{sec:theory},
then we explicit our methods of computing the unnexpectedness of past events in
cyber-physical systems~\ref{sec:computing}. We illustrate this approach by
providing an implementation relying on a few predicates on events from a smart
home simulation~(section \ref{sec:example}). In section~\ref{sec:related} we briefly
review other related works using complexity theory or trying to explain smart
homes and how our work can be linked with them before exploring potential
extensions of our work in section \ref{sec:future}

\section{Theoretical Background}
\label{sec:theory}
Kolmorov complexity formally quantifies the amount of information required
for the computation of a finite binary string\footnote{While the defintion still
holds for some inifinite binary strings (think of the representation of the
decimals of $\pi$), we restrict ourselves to finite strings in this paper.} (or
any object represented by a finite bynary
string)\cite{kolmogorov_three_1965,li_introduction_2008}. For such a binary
string $s$, its complexity $K(s)$, is the bit length of the shortest program $p$
which, if given as input to a universal Turing Machine $U$, outputs $s$.
\begin{equation}
  \label{eq:kolmo_def}
  K_{U}(s) = \min_{p}\left\{l(p)|U(p)=s\right\}
\end{equation}
The first notable property of this definition is its universality: while the
choice of the Turing machine $U$ used for the computations appears in the
definition of eq. \ref{eq:kolmo_def}, all results stand, up to an additional
constant. Think, for instance, how any Turing-complete programming language can
be turned into any other language, using a compiler program. In fact, since any
Turing machine $U'$ can be emulated by $U$ from a
finite program $p_{U}$, we have the following inequality:
\begin{equation}
  \label{eq:inequality_univ}
  K_{U}(s) \le l(p_{u}) + K_{U}(s)
\end{equation}
From this first result, we can then define a universal complexity $K(s)$, which
no longer depends on the choice of a Turing machine $U$, such that, for any
machine $U$,
\begin{equation}
  \forall U\in\text{TM}, \forall s, |K(s) - K_{u}(s)| \le C_{U}
\end{equation}
where the additional constant $C_{U}$ does not depend on the object $s$.

It is important to note that the notion of Kolmogorov complexity is not related
to the computation complexity, as there is no requirement on the execution time
of the programs, only their length in bits matters for the computation of
complexity. In fact, it can be shown that Kolmogorov complexity is not
computable\cite{li_introduction_2008}. The sketch of the proof is the following:
since the definition of complexity from eq. \ref{eq:kolm_def} requires to find
the shortest of all programs on a given Turing machine outputting $s$, it
requires to run all programs up to a given length and compare their result to
$s$. However, being able to do so means that we would be able to tell whether
these programs terminate. Since the termination of programs is famously
undecidable, by extension, the computation of Kolmogorov complexity is
impossible. One can however easily approximate it with upper bounds, by
exhibiting a program outputting $s$.

While the introducion of Kolmogorov complexity was purely theoretical, many
applications were found, notably because of its close relation with the
intuitive notions of the complexity of items, events, situations. For
instance,~\cite{murena_solving_2020} studied the use of Kolmogorov Complexity to
identify and solve analogies.

Algorithmic Information Theory is also used to define the notion of
unexpectedness: in \cite{dessalles2011coincidences}, the unexpectedness $U(e)$
of an event is defined as the difference between an a-priori expected causal
complexity $C_{w}(e)$ and the actual observed complexity $C(e)$.
\begin{equation}
  \label{eq:unexpected}
  U(e) = |C_{w}(e) - C(e)|
\end{equation}

This result comes from the understanding that, while Kolmogorov complexity is
computed with Turing machine, it can be used as an elegant proxy for modeling
information processing in the human brain, and thus helps design a notion of
simplicty or complexity of events\cite{}.

Using this measure, it is possible to model phenomenons such as coincidences:
imagine that you happen to run into a friend in a park. The unnexpectedness will
be much larger if this park is in Singapore than if it is in your common
neighbourhood: the causal complexity $C_{w}$ is much higher if you a priori have
no knowledge of your friend being around, while the description complexity
remains the same.

\section{Computing the unexpectedness of events}
\label{sec:computing}
\subsection{A memory model}
We model the memory as an unordered set $M$ of events $e$. Individual events can
be seen as data points, augmented with a label indicating their nature
(temperature events, failure events, addition/removal events), and a timestamp:
\begin{equation}
  \label{eq:event}
  e \equiv (l, t,\mathcal{D})
\end{equation}
where $l$ is the label, $t$ a timestamp and $\mathcal{D}$ a multi-dimensional
datapoint representing various characteristics of the events (e.g.~its duration,
the maximum temperature reached, the sensor name, its position).

@@

@@

From the previous definitions, we then estimate the complexity of events as the
``simplest'' way of retrieving them from the memory, by applying successive
filters. A retrieval path composed of various filters
$p = f_{1} = f_{\pi_{1}, k_{1}}, f_{2} = f_{\pi_{2}, k_{2}}, \dots, f_{n} = f_{\pi_{n}, k_{n}}$
is said to \emph{retrieve} an event $e$ from a memory $\mathcal{M}$ if and only
if
$p(\mathcal{M}) = f_{1} \circ f_{2} \circ \dots \circ f_{n} (\mathcal{M}) = \{e\}$.
For an event $e$, if such a path exist, its complexity can be bounded by
$K(p) = \sum_{f_{\pi, k} \in p} l(\pi) + l(k)$, the complexity of the retrieval
path being the sum of the bit lengths required to describe the predicates $\pi$
and their arguments $k$.

The description complexity of an event $e$ can thus be defined as the minimum complexity of a retrieval path retrieving $e$:

\begin{equation}
  \label{eq:desc_k}
  K(e) = \min_{p | p(\mathcal{M}) = e} K(p) = \min_{p | p(\mathcal{M})=e} \sum_{f_{\pi, k} \in p} l(\pi) + l(k)
\end{equation}

\begin{algorithm}
  $\mathtt{current_{explore}} \leftarrow [\mathcal{M}]$ \;
  $\mathtt{future_{explore} \leftarrow} [\;]$ \;
  $\mathtt{pass} \leftarrow 0$ \;
  \While{$\mathtt{current_{explore}} \neq [\;]$ \textbf{and} $\mathtt{pass} < \mathtt{max\_pass}$}{
    \For{$(\mathtt{M_{prev}}, \mathtt{K_{prev}}) \in \mathtt{current_{explore}}$}{
      \For{$\mathtt{p \in \mathcal{P}}$}{
        \For{$\mathtt{k \in \{0,1\}^{*}}$}{
          $\mathtt{K_{current} \leftarrow l(p) + l(k) + K_{prev}}$ \;
          \If{$\mathtt{K_{current}} > \mathtt{max_{complex}}$}{
            \textbf{break} \;
          }
          $\mathtt{M'} \leftarrow f_{p,k}(\mathtt{M_{prev}})$ \;
          \eIf{$\mathtt{M'} = \mathtt{\{e\}}$}{
            $\mathtt{K(e) \leftarrow \min(K(e), K_{current})}$ \;
          }{
            $\mathtt{future_{explore}.append((M', K_{current}))}$\;
          }
        }
      }
    }
    $\mathtt{current_{explore}} \leftarrow \mathtt{future_{explore}}$ \;
    $\mathtt{future_{explore}} \leftarrow [\;]$ \;
    $\texttt{pass} \leftarrow \mathtt{pass} + 1$\;
  }
  \caption{Iterative computation of the complexity}
  \label{alg:complex_iter}
\end{algorithm}

The definition used for description complexity in eq. \ref{eq:desc_k} allows for
a direct implemenation, which is shown in alg. \ref{alg:complex_iter}. This
algorithm operates iteratively: starting with the base memory $\mathcal{M}$
(line 1), we apply all possible predicate concepts $\pi$ and programs $k$ (lines
6-7), up to a given length $\mathtt{max\_len}$ bits, and apply them:
$M' = f_{\pi, k}(M)$ (line 12). We then store the pairs
$(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array $\mathtt{future_{explore}}$. At
the end of the iteration, the results of the filters become the memories which
will be explored during the next iteration (lines 21--23). Each pass thus explore
retrieval paths of increasing length. When a singleton memory is reached, the
complexity of its unique element is upper-bounded with the length of the
corresponding retrieval path (line 14).

@@

The term $C_{w}(e)$ from eq. \ref{eq:unexpected} is harder to formally
fit into our canvas. As $C_{w}$ is defined as the shortest program a
``W-machine'' replicating the world's behaviour, based on the person's
knowledge\cite{dessalles2011coincidences}, it is convenient to understand this
term as the expectation one would have regarding the complexity of the event
$e$, \emph{before} actually evaluating it.

@@

Now, when considering, for all possible predicates $\pi$, the corresponding
neighbourhoods $N_{\pi}(e)$, with the convention that $N_{\pi}(e) = \emptyset$
if $\pi(e) = \bot$, we can compute an average expected complexity for $e$:

@@

The definition used for the expected complexity of the event $e$ in eq.
\ref{eq:expected} holds the intuition of favorizing more similar events: if $e'$
is \emph{very} similar to $e$, it will appear in many neighbouhoods, since it
satisfies mostly the same predicates as $e$. Therefore, its complexity term will
be present in more terms in eq. \ref{eq:expected}, and will weight more towards
the final result.

@@

In our canvas of predicative filtering, this corresponds to adding for free, in
terms of program length, information about $e$ to predicates. This translates
into considering relative order between events: when investigating a cause for
an anomaly in the living-rooms, other anomalies occurring in the same
living-room will be simpler, since the location ``living-room'' is already known
from knowing the consequnce. In terms of predicates, we therefore append a
description of $e$ to all programs $k$ of predicates:
$\pi_{k}(c) = \pi_{k::e}(c)$, where $::$ is the append operation.

@@

\begin{equation}
  \label{eq:abd_k}
  Abd(c) = \min_{p | p(\mathcal{M}|e) = c} K(p) = \min_{p | p(\mathcal{M}|e)=c} \sum_{f_{\pi, k} \in p} l(\pi) + l(k)
\end{equation}
This abduction score translates the additional information provided to the
system when answering a user's request. As such, it encapsulates the idea
presented as the motivation of this paper: when confronted to a surprising
situation, and with no other source of information, it can be used as a metrics
to find relevant posisble surprising causes to the observed consequence.

\section{Implementation Example}
\label{sec:example}
\subsection{Setup}
To test our approach, we set up an experimental smart home setup. This choice of
configuration is motivated by the challenges posed by smart homes for abductive
inference: i) as the number of connected devices increases, more events are
recorded, making the detection of memorable events more important; ii) smart
homes are prone to undergo atypical situations, highly dependant on the context,
for which pre-established relations might fail to find good abduction
candidates. Furthermore, the choice was also motivated by the ease of application: as
previous works exist on smart homes, it is possible to simulate their behaviour
and quickly generate data to extract events from and test our methods.

@@

\begin{itemize}
    \item Unusual weather: the outdoor conditions are set to unusually high or
        low temeperatures.
    \item Heater failures: heater can fail, making them turning off regardless
        of the command they receive.
    \item User's vacation: the user go out of the building for an extended
        period of time.
    \item Device removal/addition: a device is removed, or another one is added
        to the system.
\end{itemize}

\begin{figure}[ht]
  \includegraphics[width=\linewidth]{figures/ts_example}
  \caption{Time series data from the simulation: outdoor temperature and
    controller temperature of a room. We can see two temperature drops, one for
    each room. What could be the causes of these drops?}
  \label{fig:ts_example}
\end{figure}

@@

For the implementation of our method, we first needed to identify and
characterize events from the time series data generated by the iCasa simulation.
Since this is not the focus point of our present work (see sec.
\ref{sec:related}), we simply apply threshold and precomputed conditions based
detections to create a base of events.

>>>>

\begin{abstract}
  When confronted to an unprecedented situation, humans typically show good
  performance in quickly identifying noticeable past events and proposing them
  as possible causal hypotheses. This kind of abductive inference is widely
  overlooked in modern AI approaches which rely on massive datasets to learn
  associative patterns. Our proposal is to formalize and compute a ``memorability''
  score over a memory of various recorded events from a cyber-physical system.
 This score can later be used either to select only relevant information to be
 remembered, or to propose causal hypotheses in unusual situations, on demand.
 As such, the approach aims at being complementary to more traditional
 learning-focused techniques. We provide theoretical ground for our approach by
 using and extending existing results and ideas from Algorithmic Information
 Theory and provide an implementation example, showing practical results in a
 smart-home scenario.
\end{abstract}

@@

As the number of connected devices and sensors grows, so does the amount of
produced data. As a consequence, the ability to filter out a few events from past recordings
as ``memorable'' becomes prime. While this task is often overlooked by many
statistical approaches, which instead benefit from large quantity of
information, it is part of the learning process for humans
\cite{dehaene_how_2020}. Faced with new situations, humans are prone to infer
previously seen unusual events as possible causes, and carry out further testing
to assess the existence of a causal relation.

For instance, it is possible, for humans without any prior knowledge about
physics or electrical engineering, to infer a memorable thunderstorm as a
possible cause for a general black-out just by using the memorability of
the former. While this kind of abductive inference (i.e. inference to the cause
of an observation \cite{magnani_abduction_2011}) can yield many
false-positives, it can be considered in situations where the lack of knowledge
from previous occurrences penalizes other approaches.

% The difficulty of this approach to reasoning is, however, to quickly identify
% the relevant candidate hypotheses. In this regards, we may use a basic
% instinct: without prior knowledge, the most relevant hypothesis might simply be
% the most memorable recent event. However this definition is highly subjective,
% and does not seem to fit well into the canvas of computing and AI.

This kind of abduction seems however hard to translate into machine, as the difficulty
comes from different factors. First, events can be of different nature, and not
directly comparable. In fact, even for specific systems such as smart homes,
events range from device removal to a presence detection or an unusually high
temperature. How can one can then assess which ones are more memorable?
Furthermore, even for comparable events, different characteristics can be put
forward to argue for the most memorable: is a record-high temperature 47 days
ago more memorable than the small deviation recorded just 3 minutes ago? To this
day, no current system proposes to use all this information from various event
types from different devices to compute a unified metrics of ``memorability''.

% The difficulty of doing so comes from the wide variety of events, parameters and
% characteristics to take into account. How can one assess if a small recent
% perturbation is more important than a larger but older one? In addition, one
% needs a common metrics to evaluate the relative importance of data coming from
% various devices. In the case of a smart home, for instance, devices range from
% presence sensors to humidity and temperature sensors, power meters, etc. To this
% day, no current system proposes to analyze all events coming from such various
% devices and come to a unified metrics of ``memorability''.

To tackle this challenge, we use as a starting point the observation that while
all events, regardless of their characteristics or nature, can be uniquely
described using a combination of qualifiers, the most memorable ones are likely
to require less words to be described. Think, for instance, of ``last year's
hottest day'' and ``the 182th day of 7 years ago''. By evaluating the lenght of
each description, taking into account both the complexity of concept words (a
time ranking, a temperature ranking), and the arguments used in both
descriptions (the hottest, the 182th, 7), we can assign each event a complexity
score. Memorable events, as they stand out, would then differ from their
neighbors by being much simpler (or more complex).

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{figures/filters}
  \caption{An example of retrieving an event through successive predicative
    filters. From the base memory (yellow), successive filters select the events
    satisfying their predicates (gray arrows). The operations yields successive
    subsets of the memory (yellow), until a final singleton memory is reach,
    meaning that the successive filters retrieve a unique event from the memory.
    The predicates used for this retrieval path and their arguments,
    $\pi_{1}, \pi_{2}, \pi_{3}, k_{1}, k_{2}, k_{3}$, can be used as a retrieval
    program. Its length gives an upper bound to the description complexity of
    the event $e_{h}$: ``last year's hottest day''.}
  \label{fig:filters}
\end{figure}

To programmatically achieve this computation, we rely on principles from
Algorithmic Information Theory, which provides formal definitions of complexity
and simplicity of objects, which appear to be consistent with human perception
\cite{li_introduction_2008, dessalles2011coincidences, delahaye_numerical_2012}.
 Figure \ref{fig:filters} shows our solution applied to the example description
 ``last year's hottest day''. We assume that
a memory $\mathcal{M}$ is a unordered set of various distinct objects, upon
which can be applied boolean predicates $\pi_{k}$ to describe them. These predicates thus
provide filters $f_{\pi, k}$ that can be applied to the memory, selecting all events
satisfying predicate $\pi_{k}$. By applying successive
filters, we narrow down the size of the memory, until a single event remains.
All the successive filters used to come to this singleton memory constitutes a
\emph{retrieval path}, whose length, in bits, gives an upper bound of the
description complexity of the retrieved event.

The rest of this paper is organized as follows: we first briefly introduce some notions of
Algorithmic Information Theory and its applications in section~\ref{sec:theory},
then we explicit our methods of computing the unexpectedness of past events in
cyber-physical systems~\ref{sec:computing}. We illustrate this approach by
providing an implementation relying on a few predicates on events from a smart
home simulation~(section \ref{sec:example}). In section~\ref{sec:related} we briefly
review other related works using complexity theory or trying to explain smart
homes and how our work can be linked with them before exploring potential
extensions of our work in section \ref{sec:future}

\section{Theoretical Background}
\label{sec:theory}
Kolmogorov complexity formally quantifies the amount of information required
for the computation of a finite binary string\footnote{While the definition still
holds for some infinite binary strings (think of the representation of the
decimals of $\pi$), we restrict ourselves to finite strings in this paper.} (or
any object represented by a finite binary
string)\cite{kolmogorov_three_1965,li_introduction_2008}. For such a binary
string $s$, its complexity $K(s)$, is the bit length of the shortest program $p$
which, if given as input to a universal Turing Machine $U$, outputs $s$.
\begin{equation}
  \label{eq:kolmo_def}
  K_{U}(s) = \min_{p}\left\{l(p)|U(p)=s\right\}
\end{equation}

The first notable property of this definition is its universality: while the
choice of the Turing machine $U$ used for the computations appears in the
definition of eq. \ref{eq:kolmo_def}, all results stand, up to an additional
constant. Think, for instance, how any Turing-complete programming language can
be turned into any other language, using a compiler program. In fact, since any
Turing machine $U'$ can be emulated by $U$ from a
finite program $p_{U}$, we have the following inequality:
\begin{equation}
  \label{eq:inequality_univ}
  K_{U}(s) \le l(p_{u}) + K_{U}(s)
\end{equation}

From this first result, we can then define a universal complexity $K(s)$, which
no longer depends on the choice of a Turing machine $U$, such that, for any
machine $U$,
\begin{equation}
  \forall U\in\text{TM}, \forall s, |K(s) - K_{u}(s)| \le C_{U}
\end{equation}
where the additional constant $C_{U}$ does not depend on the object $s$.

It is important to note that the notion of Kolmogorov complexity is not related
to computation complexity, as there is no requirement on the execution time
of the programs, only their length in bits matters for the computation of
complexity. In fact, it can be shown that Kolmogorov complexity is not
computable\cite{li_introduction_2008}. The sketch of the proof is the following:
since the definition of complexity from eq. \ref{eq:kolm_def} requires to find
the shortest of all programs on a given Turing machine outputting $s$, it
requires to run all programs up to a given length and compare their result to
$s$. However, being able to do so means that we would be able to tell whether
these programs terminate. Since the termination of programs is famously
undecidable, by extension, the computation of Kolmogorov complexity is
impossible. One can however easily approximate it with upper bounds, by
exhibiting a program outputting $s$.

Interestingly, Kolmogorov's definition of complexity matches the intuitive
notion and perception of complexity from a human standpoint. For instance, the
complexity of short binary strings evaluated in \cite{delahaye_numerical_2012}
shows similar results to human perception of complex strings and patterns. More
recently, \cite{murena_solving_2020} used Kolmogorov complexity to solve
analogies and showed results close to human expectations.

This bridge between Algorithmic Information Theory and human perception of
complexity is extended to define the notion of simplicity and unexpectedness,
which are considered of uttermost importance in cognitive science\cite{chater_simplicity_2003}.
\cite{dessalles2011coincidences} proposes a formal definition of the unexpectedness $U(e)$
of an event, as the difference between an a-priori expected causal
complexity $C_{w}(e)$ and the actual observed complexity $C(e)$.
\begin{equation}
  \label{eq:unexpected}
  U(e) = |C_{w}(e) - C(e)|
\end{equation}
This result comes from the understanding that, while Kolmogorov complexity is
computed with Turing machine, it can be used as an elegant proxy for modeling
information processing in the human brain, and thus helps design a notion of
simplicity or complexity of events.

Using this measure, it is possible to model phenomenons such as coincidences:
imagine that you happen to run into someone in a park. If this person is not
noticeable, the event will be quite trivial. On the other hand, this person is
your best friend, as the description complexity $C(e)$ drops, the causal
complexity $C_{W}(e)$ remains unchanged, and the event becomes noticeable.

Using these insights from AIT, we propose to measure the complexity of past
events $e$ from a memory $\mathcal{M}$ by considering the quantity
$K(e|\mathcal{M})$, that is the length of the shortest program outputting $e$,
given the entire memory of past events $\mathcal{M}$ as an input. To limit the
search and obtain a computable measure, we limit ourselves to only a subset of
possible programs, which we will now describe.

\section{Computing the unexpectedness of events}
\label{sec:computing}
\subsection{Retrieving an event}
We model the memory as an unordered set $M$ of events $e$. Individual events can
be seen as data points, augmented with a label indicating their nature
(temperature events, failure events, addition/removal events), and a timestamp:
\begin{equation}
  \label{eq:event}
  e = (l, t,\mathcal{D})
\end{equation}
where $l$ is the label, $t$ a timestamp and $\mathcal{D}$ a multi-dimensional
data point representing various characteristics of the events (e.g.~its duration,
the maximum temperature reached, the sensor name, its position).

@@

As the output of a filter applied to the memory $\mathcal{M}$ is another memory
object $\mathcal{M'} = f_{\pi, k}(\mathcal{M})$, we can compose filter
functions. A sequence of such filters, applied in succession, is called a
\emph{retrieval path}
\begin{equation}
\label{eq:ret_def}
p = (f_{\pi_{1}, k_{1}}, \dots, f_{\pi_{n}, k_{n}})
\end{equation}
and by definition
$p(\mathcal{M}) = f_{\pi_{n}, k_{n}}(\dots(f_{\pi_{1}, k_{1}}(\mathcal{M})))$.
In case the result of the operation $p(\mathcal{M})$ contains a single element
$e$, we say that the path $p$ \emph{retrieves} the element $e$ from $\mathcal{M}$, and write
$p(\mathcal{M}) = e$.

@@

% From the previous definitions, we then estimate the complexity of events as the
% ``simplest'' way of retrieving them from the memory, by applying successive
% filters. A retrieval path composed of various filters
% $p = f_{1} = f_{\pi_{1}, k_{1}}, f_{2} = f_{\pi_{2}, k_{2}}, \dots, f_{n} = f_{\pi_{n}, k_{n}}$
% is said to \emph{retrieve} an event $e$ from a memory $\mathcal{M}$ if and only
% if
% $p(\mathcal{M}) = f_{1} \circ f_{2} \circ \dots \circ f_{n} (\mathcal{M}) = \{e\}$.
% For an event $e$, if such a path exist, its complexity can be bounded by
% $K(p) = \sum_{f_{\pi, k} \in p} l(\pi) + l(k)$, the complexity of the retrieval
% path being the sum of the bit lengths required to describe the predicates $\pi$
% and their arguments $k$.

As presented in sec. \ref{sec:theory}, we are interested in computing the
Kolmogorov complexity of an event $e$, conditionned on the memory $\mathcal{M}$.
As this quantity is not computable, we use the previous definitions of retrieval
paths and approximate the quantity $K(e|\mathcal{M})$ by the lenght of the
shortest retrieval path $p$ retrieving $e$ from $\mathcal{M}$.

Given that all predicate concepts can be described by a prefix program of length
$l(\pi)$, and their arguments in length $l(k)$, each retrieval path $p$ can be
described by the sequence of prefix description of all the predicate concepts and
arguments used to construct its filters:

\begin{equation}
K(p) = l((f_{\pi_{1}, k_{1}}, \dots, f_{\pi{n}, k_{n}})) = \sum_{i=0}^{n} l(\pi_{i}) + l(k_{i})
\end{equation}

Using this definition of the lenght of a retrieval path, we can compute the
estimation for the complexity $K(e|\mathcal{M})$, which, for clarity, we will
denote simply by $K(e)$ for the rest of this paper.

\begin{equation}
  \label{eq:desc_k}
  K(e) \approx \min_{p | p(\mathcal{M}) = e} K(p) = \min_{p | p(\mathcal{M})=e} \sum_{f_{\pi, k} \in p} l(\pi) + l(k)
\end{equation}

\begin{algorithm}
  $\mathtt{current_{explore}} \leftarrow [(\mathcal{M}, 0)]$ \;
  $\mathtt{future_{explore} \leftarrow} [\;]$ \;
  $\mathtt{pass} \leftarrow 0$ \;
  $\mathtt{K(e)} \leftarrow +\infty$ \;
  \While{$\mathtt{current_{explore}} \neq [\;]$ \textbf{and} $\mathtt{pass} < \mathtt{max\_pass}$}{
    \For{$(\mathtt{M_{prev}}, \mathtt{K_{prev}}) \in \mathtt{current_{explore}}$}{
      \For{$\mathtt{\pi \in \mathcal{P}}$}{
        \For{$\mathtt{k \in \{0,1\}^{*}}$}{
          $\mathtt{K_{current} \leftarrow l(\pi) + l(k) + K_{prev}}$ \;
          \If{$\mathtt{K_{current}} > \mathtt{max_{complex}}$}{
            \textbf{break} \;
          }
          $\mathtt{M'} \leftarrow f_{\pi,k}(\mathtt{M_{prev}})$ \;
          \eIf{$\mathtt{M'} = \mathtt{\{e\}}$}{
            $\mathtt{K(e) \leftarrow \min(K(e), K_{current})}$ \;
          }{
            $\mathtt{future_{explore}.append((M', K_{current}))}$\;
          }
        }
      }
    }
    $\mathtt{current_{explore}} \leftarrow \mathtt{future_{explore}}$ \;
    $\mathtt{future_{explore}} \leftarrow [\;]$ \;
    $\texttt{pass} \leftarrow \mathtt{pass} + 1$\;
  }
  \caption{Iterative computation of the approximate complexity}
  \label{alg:complex_iter}
\end{algorithm}

The definition used for description complexity in eq. \ref{eq:desc_k} allows for
a direct implementation, which is shown in alg. \ref{alg:complex_iter}. This
algorithm operates iteratively: starting with the base memory $\mathcal{M}$
(line 1), we apply all possible predicate concepts $\pi$ from a given set
$\mathcal{P}$ and programs $k$ (lines 6-7), up to a given length
$\mathtt{max\_len}$ bits, and apply them: $M' = f_{\pi, k}(M)$ (line 12). We
then store the pairs $(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array
$\mathtt{future_{explore}}$. At the end of the iteration, the results of the
filters become the memories which will be explored during the next iteration
(lines 21--23). Each pass thus explore retrieval paths of increasing length.
When a singleton memory is reached, the complexity of its unique element is
upper-bounded with the length of the corresponding retrieval path (line 14).

@@

The term $C_{w}(e)$ from eq. \ref{eq:unexpected} is harder to formally
fit into our canvas. As $C_{w}$ is defined as the shortest program a
``W-machine'' replicating the world's behavior, based on the person's
knowledge\cite{dessalles2011coincidences}, it is convenient to understand this
term as the expectation one would have regarding the complexity of the event
$e$, \emph{before} actually evaluating it.

@@

Now, when considering, for all possible predicates $\pi$, the corresponding
neighborhoods $N_{\pi}(e)$, with the convention that $N_{\pi}(e) = \emptyset$
if $\pi(e) = \bot$, we can compute an average expected complexity for $e$:

@@

The definition used for the expected complexity of the event $e$ in eq.
\ref{eq:expected} holds the intuition of favoring more similar events: if $e'$
is \emph{very} similar to $e$, it will appear in many neighborhoods, since it
satisfies mostly the same predicates as $e$. Therefore, its complexity term will
be present in more terms in eq. \ref{eq:expected}, and will weight more towards
the final result.

@@

In our canvas of predicative filtering, this corresponds to adding for free, in
terms of program length, information about $e$ to predicates. This translates
into considering relative order between events: when investigating a cause for
an anomaly in the living-rooms, other anomalies occurring in the same
living-room will be simpler, since the location ``living-room'' is already known
from knowing the consequence. In terms of predicates, we therefore append a
description of $e$ to all programs $k$ of predicates:
$\pi_{k}(c) = \pi_{k::e}(c)$, where $::$ is the append operation.

@@

\begin{equation}
  \label{eq:abd_k}
  Abd(c) = \min_{p | p(\mathcal{M}|e) = c} K(p) = \min_{p | p(\mathcal{M}|e)=c} \sum_{f_{\pi, k} \in p} l(\pi) + l(k)
\end{equation}
This abduction score translates the additional information provided to the
system when answering a user's request. As such, it encapsulates the idea
presented as the motivation of this paper: when confronted to a surprising
situation, and with no other source of information, it can be used as a metrics
to find relevant possible surprising causes to the observed consequence.

\section{Implementation Example}
\label{sec:example}
\subsection{Setup}
To test our approach, we set up an experimental smart home setup. This choice of
configuration is motivated by the challenges posed by smart homes for abductive
inference: i) as the number of connected devices increases, more events are
recorded, making the detection of memorable events more important; ii) smart
homes are prone to undergo atypical situations, highly dependent on the context,
for which pre-established relations might fail to find good abduction
candidates. Furthermore, the choice was also motivated by the ease of application: as
previous works exist on smart homes, it is possible to simulate their behavior
and quickly generate data to extract events from and test our methods.

@@

\begin{itemize}
    \item Unusual weather: the outdoor conditions are set to unusually high or
        low temperatures.
    \item Heater failures: heater can fail, making them turning off regardless
        of the command they receive.
    \item User's vacation: the user go out of the building for an extended
        period of time.
    \item Device removal/addition: a device is removed, or another one is added
        to the system.
\end{itemize}

\begin{figure}[ht]
  \includegraphics[width=\linewidth]{figures/ts_example}
  \caption{Time series data from the simulation: outdoor temperature (blue) and
    controller temperature of a room (blue). On different occasion, the
    controlled temperature deviates from its setting (21°C). We will evaluate if
  our system finds these deviations as memorable, and if it can propose causal
  hypotheses.}
  \label{fig:ts_example}
\end{figure}

@@

For the implementation of our method, we first needed to identify and
characterize events from the time series data generated by the iCasa simulation.
Since this is not the focus point of our present work (see sec.
\ref{sec:related}), we simply apply threshold and pre-computed conditions based
detection to create a base of events.

This base of events consists the basis of the initial memory $\mathcal{M}$ used
for computations.
We then implemented a small number of predicate concepts:
\begin{itemize}
        \item \texttt{label(e, k)}: whether the event $e$ has the label ranked
        $k^{th}$ in the memory.
        \item \texttt{rank(e, k)}: whether the event $e$ has the rank $k_{1}$
        along axis $k_{2}$, where $k$ is a unique encoding for $(k_{1}, k_{2})$.
        \item \texttt{day(e, k)}: whether the event $e$ occurred $k$ days ago.
        \item \texttt{month(e, k)}: whether the event $e$ occurred $k$ months
        ago.
        \item \texttt{location(e, k)}: whether the event $e$ occurred in zone
        $k$.
\end{itemize}
